---
title: "hw1_sol"
author: "TA"
date: "1/9/2021"
output:
  html_document: default
  pdf_document: default
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/Documents/GitHub/ma189/Data')
```

## Q.1,2,3

*Source*: The Child Health and Development Studies (CHDS) data are presented in *Stat Labs: Mathematical Statistics Through Applications* by Deborah Nolan and Terry Speed (Springer). 



```{r}
baby <- read.table("babies.dat",header=TRUE)

head(baby[c(1,4,6)])

```
## The dataset consists of 1236 records over 7 variables.

## Q.4 Submatrix  


```{r }

X<- as.matrix(tail(baby[,c(1,4,6)], n = 5))

X

```

## Q.5

```{r}

A<- t(X)%*%X

A
```


## Q.6 
```{r}

det (A)

inv_A<- solve(A)

inv_A
```
## The Determinant of A is not zero, hence the inverse exists.



## Q.7 

```{r}

sum(diag(A))

```
## The trace of the matrix is positive which means the sum of it's eigenvalues is also positive as trace(A) = sum of it's eigenvalues.

## Q.8 
One way to check if a matrix is positive definite:
A matrix is positive definite if it is symmetric and all it's eigenvalues are positive 

```{r}

all(t(A) == A)

eig <- eigen(A)

eig_values <- eig$values

eig_values


```

We can see that the matrix is symmetric and all it's eigenvalues are positive. Therefor it is a positive definite matrix.