---
title: "hw2_sol"
author: "TA"
date: "1/9/2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/Documents/GitHub/ma189/Data')
```

## Q.1

Source: The file contains data extracted from the 1974 Motor Trend
US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32
automobiles (1973–74 models). The dataset contains 32 observations on 11 variables.

```{r}
mtcars <- read.table("mtcars.csv",sep = ",", header=TRUE)

head(mtcars)
# colMeans(mtcars[,2:11])


# apply(mtcars[,2:11], 2, var)

typeof(mtcars)

plot(x = mtcars$wt, y = mtcars$mpg,
     xlab = "Weight", ylab = "Miles Per Gallon",	 
     main = "Weight vs. miles per gallon")
```
## The scatter plot shows a decreasing linear relationship between mpg and weight which can be investigated further if the relationship is statistically significant.

## Q. 2
```{r}

library("scatterplot3d")
scatterplot3d(x = mtcars$wt, y = mtcars$mpg, z=mtcars$cyl,
              xlab = "Weight", ylab = "Miles Per Gallon",	zlab="Cylinders", 
              color="red", pch=20)
```
### One of the drawback of 3d scatter-plot is that the interpretation of relationships between the variables can be difficult to observe. An alternative can be to plot a 2d catter plot b/w mpg and weight and colorcode the points according to cylinder variable.


## Q. 3

```{r}

pairs(mtcars[,2:11], pch=20)

```
## Q.4

If we look at the 2d scatter plot, and color code the points accoring to the number of cylinders, we can see that the points cluster into 3 groups according to the number of cylinders. Therefore, the engineer's suggestions reasonable.
```{r}
library(ggplot2)
ggplot(mtcars, aes(x = wt, y = mpg)) +
    geom_point(aes(color = factor(cyl)))
```



