---
title: 'Math 189: Time Series III'
output: html_document
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/Documents/GitHub/ma189/Data')
```

# Linear Filters

- A *linear filter* maps an input time series $\{ X_t \}$ to an output time series $\{ Y_t \}$ by taking a linear combination of past, present, and future observations:
\[
  Y_t = \sum_{k \in {\mathbb Z}} \psi_k \, X_{t-k}.
\]
 The *filter coefficients* (or *filter weights*) are $\{ \psi_k \}$. Note the convention that $\psi_k$ weights an observation occuring $k$ time points in the past (viewing time $t$ as the present). 
 
## Example:  Simple Moving Average

- In nonparametric regression we may estimate a slowly-changing mean via averaging over neighboring values, e.g.,
\[
   \frac{1}{2m+1} \sum_{k = -m}^m \psi_k \, X_{t-k}.
\]
 This weights the past $m$ and the future $m$ observations equally, and is an example of a linear filter.  It also called a *moving average*, because the observations are averaged (over a window of $2m+1$ time points) in a way that moves over the time series. A *simple moving average* has equal weights (a general moving average could have unequal weights).
 
## Example: Smoothers Applied to Gasoline Sales

- Linear filters that suppress oscillations and reveal long-term trends are called *smoothers*. A simple moving average is an example of a smoother.
- We apply a simple moving average with $m=10$ to the logged seasonally adjusted gasoline series.

```{r}
gassa <- read.table("GasSA_2-11-13.dat")
gassa.log <- ts(log(gassa),start=1992,frequency=12)

h <- 10
simple.ma <- rep(1,2*h+1)/(2*h+1)
gas.trend <- filter(gassa.log,simple.ma,method="convolution",sides=2)
gas.trend <- ts(gas.trend,start=1992,frequency=12)

### Movie
delay <- 0
flex <- 0
siglen <- length(gas.trend) - 2*h
range <- seq(1,siglen,1)
for(t in range)
{
	Sys.sleep(delay)
	filterweight <- c(rep(NA,t-1),simple.ma + gas.trend[h+t] + flex,
		rep(NA,siglen-t))
	plot(ts(gassa.log,start=1992,frequency=12),col=1,ylab="",xlab="Year",lwd=2)
	lines(ts(gas.trend[1:(h+t)],start=1992,frequency=12),col=2,lwd=2)
	lines(ts(filterweight,start=1992,frequency=12),col="#00FF0090",lwd=2)
}
```

## Paradigm:  Difference Filter

- To suppress long-term dynamics, we can *difference*: 
\[
   Y_t = X_t - X_{t-1}.
\]
 This is a filter with weights $\psi_0 = 1$, $\psi_1 = -1$, and zero otherwise. It is called the *differencing filter*. This filter reduces polynomials in time by one degree; lines are reduced to constants.
 
## Example: Differencing Applied to Gasoline Sales

- We apply differencing to logged seasonally adjusted gasoline sales. The result can be interpreted as a *growth rate*. 

```{r}
gas.diff <- diff(gassa.log)
par(oma=c(2,0,0,0),mar=c(2,4,2,2)+0.1,mfrow=c(2,1),cex.lab=.8)
plot(gassa.log,xlab="",ylab="Log Gas  Sales",yaxt="n",xaxt="n")
axis(1,cex.axis=.5)
axis(2,cex.axis=.5)
plot(gas.diff,xlab="",ylab = "Gas Growth Rate",yaxt="n",xaxt="n")
axis(1,cex.axis=.5)
axis(2,cex.axis=.5)
mtext(text="Year",side=1,line=1,outer=TRUE)
```

## The Backward Shift

- We define the *backward shift* (or *lag*) operator $B$, which shifts a time series back one time unit. This is a linear filter that lags time by one unit:
\[
    Y_t = X_{t-1}.
\]
 So the filter weights are $\psi_1 = 1$ and zero otherwise. Informally we write $B X_t = X_{t-1}$.
 
## Powers of Backward Shift

- Then $X_{t-k} = B^k X_t$, and a general filter is expressed as
\[
  Y_t = \sum_{k \in {\mathbb Z}} \psi_k \, X_{t-k}
   =  \sum_{k \in {\mathbb Z}} \psi_k B^k \, X_{t},
\]
 so we write the filter as  $\sum_{k \in {\mathbb Z}} \psi_k B^k$. Call this $\Psi (B)$. (Mathematically, $\Psi (z)$ is a Laurent series.)
 
### Simple Moving Average Filter

- Expressed as $\Psi (B) = {(2m+1)} \sum_{k = -m}^m B^k$.

## Differencing Filter

- Expressed as $\Psi (B) = 1 -B$.
  
# Trends

- Trends measure the long-term behavior of a time series.
- Trends are steady movements in the level of a time series.
- We can use smoothers to estimate trends.
- Removing a trend by subtraction allows us to see other features.

## Example: Western Housing Starts Cycle

- We estimate and eliminate the trend in the Westing Housing Starts data, so that we can more clearly see the business cycle movements.

```{r}
hpsa <- function(n,period,q,r)
{
	# hpsa
	#     gives an HP filter for seasonal data
	#	presumes trend+seas+irreg structure
	#		trend is integrated rw
	#		seas is seasonal rw
	#		irreg is wn
	#	q is snr for trend to irreg
	#	r is snr for seas to irreg

# define trend differencing matrix

delta.mat <- diag(n)
temp.mat <- 0*diag(n)
temp.mat[-1,-n] <- -2*diag(n-1)
delta.mat <- delta.mat + temp.mat
temp.mat <- 0*diag(n)
temp.mat[c(-1,-2),c(-n,-n+1)] <- 1*diag(n-2)
delta.mat <- delta.mat + temp.mat
diff.mat <- delta.mat[3:n,]

# define seasonal differencing matrix

delta.mat <- diag(n)
temp.mat <- 0*diag(n)
inds <- 0
for(t in 1:(period-1))
{
	temp.mat <- 0*diag(n)
	temp.mat[-(1+inds),-(n-inds)] <- 1*diag(n-t)
	delta.mat <- delta.mat + temp.mat
	inds <- c(inds,t)
}
sum.mat <- delta.mat[period:n,]

# define two-comp sig ex matrices

#trend.mat <- solve(diag(n) + t(diff.mat) %*% diff.mat/q)
#seas.mat <- solve(diag(n) + t(sum.mat) %*% sum.mat/r)
trend.mat <- diag(n) - t(diff.mat) %*% solve(q*diag(n-2) + diff.mat %*% 
	t(diff.mat)) %*% diff.mat
seas.mat <- diag(n) - t(sum.mat) %*% solve(r*diag(n-period+1) + sum.mat %*% 
	t(sum.mat)) %*% sum.mat

# define three-comp sig ex matrices

trend.filter <- solve(diag(n) - trend.mat %*% seas.mat) %*%
	trend.mat %*% (diag(n) - seas.mat)
seas.filter <- solve(diag(n) - seas.mat %*% trend.mat) %*%
	seas.mat %*% (diag(n) - trend.mat)
irreg.filter <- diag(n) - (trend.filter + seas.filter)

filters <- list(trend.filter,seas.filter,irreg.filter)
return(filters)
}

Wstarts <- read.table("Wstarts.b1",skip=2)[,2]
Wstarts <- ts(Wstarts,start = 1964,frequency=12)
n <- length(Wstarts)
q <- .0001
r <- 1
hp.filters <- hpsa(n,12,q,r)

wstarts.trend <- hp.filters[[1]] %*% Wstarts
wstarts.seas <- hp.filters[[2]] %*% Wstarts
wstarts.cycle <- hp.filters[[3]] %*% Wstarts
wstarts.sa <- wstarts.trend + wstarts.cycle

comps <- ts(cbind(wstarts.trend,wstarts.seas,wstarts.cycle),start=1964,frequency=12)
trend <- ts(wstarts.trend,start=1964,frequency=12)
cycle <- ts(wstarts.cycle,start=1964,frequency=12)

par(oma=c(2,0,0,0),mar=c(2,4,2,2)+0.1,mfrow=c(3,1),cex.lab=.8)
plot(Wstarts, ylab="West Starts",xlab="",yaxt="n",xaxt="n")
axis(1,cex.axis=.5)
axis(2,cex.axis=.5)
plot(trend,xlab="",ylab = "Trend",yaxt="n",xaxt="n")
axis(1,cex.axis=.5)
axis(2,cex.axis=.5)
plot(cycle,xlab="",ylab = "Business Cycle",yaxt="n",xaxt="n")
axis(1,cex.axis=.5)
axis(2,cex.axis=.5)
mtext(text="Year",side=1,line=1,outer=TRUE)
```

## Paradigm: Nonparametric Trend Estimation

- We consider trends $\mu_t$ to be either a deterministic (but unknown) function of time, or are an unobserved stochastic process:
\[
  X_t = \mu_t + Z_t.
\]
 Here ${\mathbb E} [ X_t ] = \mu_t$ and $\{ Z_t \}$ is a mean zero time series.
- A *two-sided moving average* is a linear filter that can be used to estimate trends:
\[
  \Psi (B) = \sum_{k = -m}^m \psi_k B^k.
\]
 When all the weights are equal, this is a simple moving average. It is two-sided because the moving average uses past and future data:
 \[
  \widehat{\mu}_t = \Psi (B)  X_t = \sum_{k= -m}^m \psi_k X_{t-k}.
\]
- A *symmetric moving average* weights future and past equally, i.e., $\psi_k = \psi_{-k}$.
- Because the filter uses $m$ past and $m$ future observations, we can only compute trend estimates for times $m+1 \leq t \leq T-m$, where the sample consists of times $1,2, \ldots,T$.
- As an estimator $\widehat{\mu}_t$ of the unknown trend $\mu_t$, there is less bias if $m$ is small, and less variance if $m$ is large. This is a **Bias-Variance Dilemma**.
- Example with US population, and $m = 10$ for a simple moving average.

```{r}
pop <- read.table("USpop.dat")
pop <- ts(pop, start = 1901)

p <- 10
mu.smooth <- filter(pop,rep(1,2*p+1)/(2*p+1),method="convolution")
mu.smooth <- ts(mu.smooth,start= 1901)

par(mar=c(4,4,2,2)+0.1,cex.lab=.8)
plot(ts(pop,start=1901),lwd=3,lty=1,col=gray(.8),
	ylab="U.S. Population",xlab="Year",yaxt="n",xaxt="n")
axis(1,cex.axis=.5)
axis(2,cex.axis=.5)
lines(ts(mu.smooth,start=1901),lwd=1,lty=3,col=1)
```

## Paradigm: Trend Elimination

- We can remove a trend by subtracting off its estimate.
- We can also apply the differencing filter.
- Example with US population, detrended by a $m = 10$ simple moving average and by the difference filter.

```{r}
pop.diff <- diff(pop)

par(mar=c(4,4,2,2)+0.1,cex.lab=.8)
plot(ts(pop - mu.smooth,start=1901),lwd=1,lty=3,ylim=c(-3.5*10^6,3.5*10^6),
	ylab="U.S. Population",xlab="Year",yaxt="n",xaxt="n")
axis(1,cex.axis=.5)
axis(2,cex.axis=.5)
lines(pop.diff,col=gray(.8),lwd=2)
```
  