---
title: 'Math 189: Logistic Regression II'
output: html_document
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/Documents/GitHub/ma189/Data')
```

# Multiple Logistic Regression

- Goal: predict a binary response using multiple predictors.
- We generalize the previous model as follows:
\[
 \log \left( \frac{ p( \underline{X})}{ 1- p (\underline{X})} \right)
 = \beta_0 + \beta_1 X_1 + \ldots + \beta_p X_p
\]
where is $\underline{X}$ is a $p$-vector of covariates.
- Equivalently,
\[
 p( \underline{X}) = \frac{ \exp \{  \beta_0 + \beta_1 X_1 + \ldots + \beta_p X_p \} }{
  1 + \exp \{  \beta_0 + \beta_1 X_1 + \ldots + \beta_p X_p \} }
\]
- Let $\underline{\beta} = {[ \beta_1, \ldots, \beta_p ]}^{\prime}$ be the vector of regression coefficients; $\beta_0$ is referred to as the intercept.

## Logistic Function

- Set the logistic function to be
\[
 \phi (t) = \frac{1}{1 + e^{-t}}
\]
- Then the logistic regression model for ${(Y,\underline{X})}$ is
\[
 p( \underline{x}) = {\mathbb P} [ Y = 1 \vert \underline{X} = \underline{x} ] 
  = \phi ( \beta_0 + \underline{x}^{\prime} \underline{\beta})
\]
- On the other hand
\[
 {\mathbb P} [ Y = 0 \vert \underline{X} = \underline{x} ] 
  = 1 - \phi ( \beta_0 + \underline{x}^{\prime} \underline{\beta})
\]
 
```{r}
x <- seq(-10,10,.1)
y <- (1+exp(-x))^{-1}
plot(ts(y,start=-10,frequency=10),xlab="x",ylab="")
```

## Likelihood Function

- To compute the likelihood function, assume that $(Y_1, \underline{X}_1),\ldots,(Y_n,\underline{X}_n)$ are i.i.d. 
- The conditional density of $Y_i$ given $\underline{X}_i$ is
\[
  { \left( \phi ( \beta_0 + \underline{X}_i^{\prime} \underline{\beta} ) \right)}^{Y_i}
  \,  { \left( 1 - \phi ( \beta_0 + \underline{X}_i^{\prime} 
  \underline{\beta} ) \right)}^{1 - Y_i}
\]
Hence, the joint density of $Y_1, \ldots, Y_n$ given $\underline{X}_1, \ldots,\underline{X}_n$ is
\[
 L_n (\beta_0, \underline{\beta} ) = \prod_{i=1}^n  { \left( \phi ( \beta_0 + \underline{X}_i^{\prime} \underline{\beta} ) \right)}^{Y_i}
  \,  { \left( 1 - \phi ( \beta_0 + \underline{X}_i^{\prime} 
  \underline{\beta} ) \right)}^{1 - Y_i}
\]
This $L_n: {\mathbb R}^{p+1} \rightarrow [0,\infty)$ is called the *likelihood function*.

## Maximum Likelihood Estimator

- The *maximum likelihood estimator*  $(\widehat{\beta}_0, \widehat{\underline{\beta}})$
is defined as the maximizer of the likelihood function:
\[
(\widehat{\beta}_0, \widehat{\underline{\beta}}) = \arg \max 
L_n (\beta_0, \underline{\beta} ) 
\]
- In both theory and practice, it is easier to work with the logarithm of the likelihood function.
\[
 \mathcal{L}_n (\beta_0, \underline{\beta} ) = \log L_n (\beta_0, \underline{\beta} )
 =   \sum_{i=1}^n \left\{ Y_i \, \log \left( \phi ( \beta_0 + \underline{X}_i^{\prime} \underline{\beta} ) \right) +  (1 - Y_i) \log \left( 1 - \phi ( \beta_0 + \underline{X}_i^{\prime}   \underline{\beta} ) \right) \right\}
\]
 
### Example: Default Data

- Consider Default data. Let $X_1$ be balance, $X_2$ be income, $X_3 \in \{ 0, 1 \}$ is student status. 
 
```{r}
library(ISLR)
data(Default)
y <- as.numeric(Default$default=="Yes")
stu <- as.numeric(Default$student=="Yes")
income <- 0.001*Default$income
balance <- Default$balance
all.fit <- glm(y~stu+balance+income,data=Default,family=binomial)
summary(all.fit)
```

- We can make predictions by plugging in the coefficient estimates:
\[
 \widehat{p}( \underline{x})  =
  \phi ( \widehat{\beta}_0 + \underline{x}^{\prime} \widehat{\underline{\beta}})
\]

```{r}
pred_all <- function(obs){
  x <- c(1,obs)
  pred <- as.numeric(as.numeric(x %*% all.fit$coefficients))
  pred <- 1/(1+exp(-pred))
  return(pred)
}
```

- A student with a credit card balance of 1500 dollars and an income of 40,000 has an estimated probability of default of `r pred_all(c(1,1500,40))`.  The same balances for a non-student yield a default risk of `r pred_all(c(0,1500,40))`.
- The two-class logistic regression models have multiple-class extensions (*multiple-class logistic regression*), but in practice they tend not to be used all that often. The software for it is available in R.
- We can generate additional plots: probability of default as a function of Balance, for student (orange) and non-student (blue) groups. Horizontal line is the straight proportion of defaults for student and non-students, ignoring Balance.

```{r}
fitted.prob <- as.numeric(all.fit$fitted.values)
prob.stu <- fitted.prob[stu==1]
balance.stu <- balance[stu==1]
M1 <- matrix(c(prob.stu,balance.stu), ncol = 2)
M1 <- M1[order(M1[,1], decreasing = FALSE),]

plot(M1[,2], M1[,1],type='l',lwd=2,col="orange",
     xlab="Credit Card Balance",ylab="Default Rate")

prob.nonstu <- fitted.prob[stu==0]
balance.nonstu <- balance[stu==0]
M2 <- matrix(c(prob.nonstu,balance.nonstu), ncol = 2)
M2 <- M2[order(M2[,1], decreasing = FALSE),]

lines(M2[,2],M2[,1],lwd=2,col="slateblue1")

stu.rate <- sum(stu[y==1])/sum(stu)
nonstu.rate <- sum((1-stu)[y==1])/sum(1-stu)

abline(h=stu.rate,lwd=2,lty=2,col="orange")
abline(h=nonstu.rate,lwd=2,lty=2,col="slateblue1")
```

- We can examine boxplots for Balance as a function of Student status. 

```{r}
boxplot(balance~Default$student,xlab="Student Status", 
        ylab="Credit Card Balance",col=c("slateblue1","orange"))
```
