---
title: "Math 189: Exploratory Data Analysis I"
output: html_notebook
---
 
# Challenges in Multivariate Data Analysis

## Objectives: 
What is the scientific or social problem of interest? How can it be parsed as a multivariate analysis problem? 

## Dataset: 
Data is noisy (it has *randomness*). 

- Measurement error.  (Human or machine.) 
- Missing data. Corruption of data.
- Normality versus extreme-value.
- Homogeneity vs Heterogeneity.

## Methodology:
Which method should we use? What are the assumptions, and are they met? Are the method statistically accurate? Are they omputationally efficient?

# Exploratory Data Analysis 

An important step in data analysis is exploratory data analysis (EDA).

EDA methods include loading, cleaning, and visualizing the data; descriptive statistics are used to provide quick summaries.

## Benefits of EDA:

1. Check the quality of data: has it been cleaned?  Are there missing data or outliers?
2. Gain a first impression of data: data type, distribution, symmetry, etc.
3. Illustrate analysis results: gain intuition, collaboration, general audience, etc.

# Visualization of Multivariate Data

Why do we look at graphical displays of the data? We need to condense the data (without losing important features) so that we can digest it.

## Visualization may:

1. suggest a plausible model for the data,
2. assess validity of model assumptions,
3. detect outliers or suggest a plausible normalizing transformation

# Scatter Plot

A scatter plot is a data visualization tool that uses dots to represent the values obtained for two different variables.

## Plotted on Cartesian coordinates: 
x-axis is the value of the first variable and y-axis is the value of the second variable.

Used to check the relationship between two variables:

1. Correlation
2. Linear or nonlinear
3. Joint normality
4. Groups

## Example 1:USDA Women's Health Survey

- In 1985, the USDA commissioned a study of women’s nutrition. Nutrient intake was measured for a random sample of 737 women aged 25-50 years.
- The following variables were measured:

1. Calcium (mg)
2. Iron (mg)
3. Protein (g)
4. Vitamin A (μg)
5. Vitamin C (mg)

```{r}
nutrient <- read.table("C:\\Users\\neide\\Documents\\GitHub\\ma189\\Data\\nutrient.txt")
nutrient$V1=NULL
colnames(nutrient)=c("Calcium", "Iron", "Protein", "Vitamin A", "Vitamin C")
dim(nutrient)
head(nutrient)
```

### Scatter Plot of Calcium Versus Iron

```{r}
plot(x = nutrient$Calcium, y = nutrient$Iron,
     xlab = "Calcium", ylab = "Iron",	 
     main = "Calcium vs Iron")
points(nutrient$Calcium[5],nutrient$Iron[5],col=2,pch=19)
```

Each circle is an observation; the fifth observation is in red.

### Questions About this Plot

- Is there a structure to this plot?
- Are there outliers?
- What model should we use to fit the data?
- Can we model the data using the bivariate normal distribution?

## Data Standardization

- Rescale data from different sources and measures to a "standard" scale
- Avoid comparing apples to oranges
- A common standardization is called the *Z-score* scaling, which scales a random sample to have zero sample mean and unit sample variance.
- For a random sample $x_1, \ldots, x_n$, the Z-score scaling transforms each observation by
\[
 z_i = \frac{x_i - \overline{x}}{s}
\]
where $\overline{x}$ is the **sample mean** and $s$ is **sample standard deviation**.

```{r}
x <- nutrient[,c(2,3)]
z <- x
for(i in 1:2)
{
  z[,i] <- (x[,i] - mean(x[,i]))/sd(x[,i])
}
plot(z[,1],z[,2],xlab=names(nutrient)[2],ylab=names(nutrient)[3],
          main = "Calcium vs Iron (Standardized)")
abline(h=3,col=4)
abline(v=3,col=4)
```

We expect $99.7 \%$ of normal data to be within 3 standard deviations of the mean, or in lower left quadrant given by blue lines; here there are more than expected outliers.

```{r}
length(intersect(seq(1,dim(z)[1])[z[,1] < 3],seq(1,dim(z)[1])[z[2]< 3]))/dim(z)[1]
```

## Transformation Methods

Sometimes data is "irregular”: non-normal, outliers, skewed, heavy-tailed, etc.

- Data transformation techniques can be used to stabilize variance, so that data more resembles the normal distribution.

- Power transformation:
\[
 y = x^{\alpha} \quad 0 < \alpha < 1
\]

- Log transformation:
\[
  y = \log(x), \; 0 < x
\]
 
### Scatter Plot after Log Transformation

```{r}
x <- log(nutrient[,c(2,3)])
plot(x[,1],x[,2],xlab=names(nutrient)[2],ylab=names(nutrient)[3],
          main = "Calcium vs Iron (Log Transformation)")
```

Less outliers, reduced variance, more normal.


## Scatter Plot for Three Variables

The scatter plot can be extended to visualize the relationship among three different variables, which is called 3d scatter plot.

Plotted on Cartesian coordinates: 

1. x-axis is the value of the first variable
2. y-axis is the value of the second variable
3. z-axis is the value of the third variable

### Calcium, Iron and Protein

- Each red dot represents an observation
1. Calcium on x-axis
2. Iron on y-axis
3. Protein on z-axis

```{r}
library("scatterplot3d")
scatterplot3d(x = nutrient$Calcium, y = nutrient$Iron, z=nutrient$Protein,
              xlab = "Calcium", ylab = "Iron",	zlab="Protein", 
              color="red", pch=20)
```

### 3D Scatter Plot After Log Transformation

- Each blue represents an observation
- More clustered in a “ball” or “ellipse”
- Reduced variance
- Less outliers


```{r}
scatterplot3d(x = log(nutrient$Calcium), y = log(nutrient$Iron), z=log(nutrient$Protein),
              xlab = "Calcium", ylab = "Iron",	zlab="Protein", 
              color="steelblue", pch=20)       
```

### Pros and Cons of 3D Scatter Plot

#### Pros
- Visualization for 3 variables (can be extended to 4)
- Complex relationship rather than pair wise
- Joint sample distribution

#### Cons
- Hard to see unless you rotate it (angle dependent)
- Hard to interpret
- Cannot visualize 5 or more variables

## Pairwise Scatter Plot for More Variables
- The pairwise scatter plot aims to visualize the relationship for each pair of variables in a multivariate dataset.
- A **pairwise scatter plot** is an array of scatter plots, the $(i,j)$th plot in the array is the scatter plot between the $i$th and $j$th variables.


```{r}
pairs(nutrient, pch=20)
```

### Pros and Cons of Pairwise Scatter Plot

#### Pros
- Visualization for many variables simultaneously
- Interpret pairwise relationships 

#### Cons
- No joint relationship for more than 2 variables
- Huge array when the number of variables is large

# Time Plot

A time plot (sometimes called a *time series graph*) displays values versus time. It is similar to a scatter plot, but x-axis is chosen to be time.  

- A time plot is useful to compare the "growth” of multiple variables with respect to time (or some other common index).

- Application of time plot:

1. Finance: compare multiple stock returns versus time
2. Clinical: compare multiple patients versus survival time
3. Biology: compare multiple genome sequences versus positions
4. Physics: multiple measurements versus time
5. Economoics: examine unemployment rate and economic indicators over time

## Time Plot for Financial Time Series 

*Time series* are one of the most common data types encountered in finance, economics, and weather forecasting

- Visualization tool: time plot for multiple assets.

- Things to check in a time plot:

1. Co-movement of variables
2. Trend (increasing or decreasing)
3. Periodical patterns (daily, weekly, annual, etc.)
4. **Black swan** event (huge gap, financial crisis, etc.)

## Example 2: Stock Prices of High-tech Companies

- We collect daily stock prices (closing price) of four leading high-tech companies between January 2018 and January 2019.

- The following variables were included:

1. Daily stock price of Apple
2. Daily stock price of Facebook
3. Daily stock price of IBM
4. Daily stock price of Microsoft



<!-- # SCRAP -->

<!-- ## Stock data -->

<!-- stock=read.table("https://www.math.ucsd.edu/~wez243/data/Stock.csv", sep = ",", header=TRUE) -->
<!-- stock$index = as.numeric(row.names(stock)) -->
<!-- stock=stock[order(stock$index),] -->

<!-- png(file = "Stock_1.png", width=640, height=480) -->

<!-- # Multivariate timeseries plot of stock prices. -->
<!-- plot(x = seq(1,250), y = stock$Apple, -->
<!--      xlab = "Date", ylab = "Stock Price", ylim=c(80,240), xaxt='n',	  -->
<!--      main = "Daily Stock Prices in 2018", -->
<!--      col="red",  type="l", lwd=2) -->

<!-- lines(x = seq(1,250), y = stock$Facebook, col="blue",  lwd=2) -->

<!-- lines(x = seq(1,250), y = stock$IBM, col="black",  lwd=2) -->

<!-- lines(x = seq(1,250), y = stock$Microsoft, col="purple",  lwd=2) -->

<!-- axis(1, at = c( 50, 100, 150, 200 ), labels = c("3/27", "6/7", "8/17", "10-29")) -->

<!-- legend("topleft",, c("Apple", "Facebook", "IBM", "Microsoft"),  -->
<!--        col=c("red", "blue", "black", "purple"), lty=c(1,1,1,1)) -->

<!-- # Save the file. -->
<!-- dev.off() -->


<!-- # Calculate Daily log return -->

<!-- Apple_return=diff(log(stock$Apple), lag=1) -->
<!-- Facebook_return=diff(log(stock$Facebook), lag=1) -->
<!-- IBM_return=diff(log(stock$IBM), lag=1) -->
<!-- Microsoft_return=diff(log(stock$Microsoft), lag=1) -->

<!-- # Multivariate timeseries plot of stock returns. -->
<!-- png(file = "Stock_2.png", width=640, height=480) -->

<!-- plot(x = seq(1,249), y = Apple_return, -->
<!--      xlab = "Date", ylab = "Stock Price",  ylim=c(-0.2, 0.1), xaxt='n',	  -->
<!--      main = "Daily Stock Returns in 2018", -->
<!--      col="red",  type="l", lwd=1) -->

<!-- lines(x = seq(1,249), y = Facebook_return, col="blue",  lwd=1) -->

<!-- lines(x = seq(1,249), y = IBM_return, col="black",  lwd=1) -->

<!-- lines(x = seq(1,249), y = Microsoft_return, col="purple",  lwd=1) -->

<!-- axis(1, at = c( 50, 100, 150, 200 ), labels = c("3/27", "6/7", "8/17", "10-29")) -->

<!-- legend("bottomleft",, c("Apple", "Facebook", "IBM", "Microsoft"),  -->
<!--        col=c("red", "blue", "black", "purple"), lty=c(1,1,1,1)) -->

<!-- # Save the file. -->
<!-- dev.off() -->

<!-- png(file = "p5-2.png", width=640, height=480) -->
<!-- pairs(nutrient_std, pch=20) -->
<!-- dev.off() -->

