---
title: "Math 189: Matrix Algebra"
output: html_notebook
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/neide/Documents/GitHub/ma189/Data')
```

# I. Multivariate Data

## Matrix Formulation
In multivariate analysis, we are interested in jointly analyzing multiple dependent variables.

- These variables can be represented using matrices and vectors.

- The benefits of matrix form:

1. Simplicity in notation
2. Expressing important formulas in readable format

## Example 1: Multivariate Data of Infant Death
Data records several variables for a baby at birth:

1. Baby Weight (**bwt**): in ounces
2. Gestation (**gestation**): in days

Rows are babies, columns are variables

```{r}
baby <- read.table("babies.dat",header=TRUE)
head(baby[,c(1,2)])
```

**Notice anything weird?!**

*Source*: The Child Health and Development Studies (CHDS) data are presented in *Stat Labs: Mathematical Statistics Through Applications* by Deborah Nolan and Terry Speed (Springer).  

## Math Notation
Let $\underline{x} = [ x_1, x_2 ]'$ denote the <span style="color: red;">column</span> vector of **bwt** and **gestation**.
\[
 \underline{x} = \left[ \begin{array}{c} x_1 \\ x_2 \end{array} \right]
 = \left[ \begin{array}{c} \mbox{bwt} \\ \mbox{gestation} \end{array} \right]. 
\]
 So $\underline{x} \in {\mathbb R}^2$.
 
 There are $n=1236$ babies. Let $\underline{x}_i$ for $1 \leq i \leq 1236$ denote both variables for baby $i$, i.e.,
\[
  \underline{x}_1  = \left[ \begin{array}{c} 120  \\ 284 \end{array} \right], \quad
  \underline{x}_2  = \left[ \begin{array}{c} 113  \\ 282 \end{array} \right], \quad
    \underline{x}_3  = \left[ \begin{array}{c} 128  \\ 279 \end{array} \right],\, \cdots 
\]
 This is called the **record** for the $i$th baby (or unit). Let $i$ index baby, and $j = 1, 2$ index variable. Then $x_{i,j}$ is **bwt** ($j=1$) or **gestation** ($j=2$) for baby $i$. Gestation for the fourth baby is $x_{4,2} =$ `r baby[4,2]`. 
 
 A data matrix consists of all **records** as <span style="color: blue;">row</span> vectors:
\[
  \left[ \begin{array}{c} \underline{x}_1^{\prime}  \\ \underline{x}_2^{\prime} \\ \underline{x}_3^{\prime} \end{array} \right]
  = \left[ \begin{array}{cc} 120 & 284 \\ 113 & 282  \\ 128  & 279  \end{array} \right]. 
\]

```{r}
x <- baby[1:3,1:2]
x
```

Each <span style="color: blue;">row</span> is information for one baby; each <span style="color: red;">column</span> is all measurements (across babies) of either **bwt** or **gestation**

## Definition of Matrix and Vector
A matrix is a two-dimensional array of numbers, symbols, or expressions, arranged in rows and columns.

- A vector is a matrix with either one row or one column.

- The dimension of a matrix is two numbers: number of <span style="color: blue;">rows</span> (units) and number of <span style="color: red;">columns</span> (units). Dimension of baby is `r dim(baby)` and dimension of submatrix $x$ is `r dim(x)`.

- A square matrix is one for which the numbers of <span style="color: blue;">rows</span> and number of <span style="color: red;">columns</span> are equal, e.g.

```{r}
x <- baby[1:2,1:2]
dim(x)
```
 
## Notations of Matrix and Vector

- Usually in this course we only consider matrices and vectors whose entries are real numbers.

- A real-valued **matrix** ${\bf M}$ of dimension $n \times m$ is denoted ${\bf M} \in {\mathbb R}^{n \times m}$.

- A real **vector** $\underline{v}$ of length $n$ is denoted $\underline{v} \in {\mathbb R}^n$. 

- A <span style="color: blue;">row</span> vector can be written $\underline{v} \in {\mathbb R}^{1 \times n}$, and a <span style="color: red;">column</span> vector can be written $\underline{v} \in {\mathbb R}^{n \times 1}$.

- A **scalar** is a real number, or a vector of length $1$, or a $1 \times 1$ dimensional matrix.

```{r}
x <- as.matrix(baby[1:3,1:2])
x
x[,1]
x[,1,drop=FALSE]
x[1,1]
x[1,1,drop=FALSE]
```
 
## Transpose of a Matrix

- Definition: flip the rows and columns by interchanging row and column index. Denoted by ${\bf M}^{\prime}$.

- If ${\bf M}_{i,j}$ is <span style="color: blue;">row</span> $i$, <span style="color: red;">column</span> $j$ of matrix ${\bf M}$, then ${\bf M}^{\prime}_{i,j} = {\bf M}_{j,i}$, so that now $i$ is the <span style="color: red;">column</span> index, and $j$ is the <span style="color: blue;">row</span> index.
 
```{r}
x
t(x)
```

 
 
## Symmetric Matrices

- Definition: A square matrix ${\bf M}$ is symmetric if ${\bf M}^{\prime} = {\bf M}$.

- Hence ${\bf M}_{i,j} = {\bf M}_{j,i}$.

```{r}
m <- t(x) %*% x
m
```

- Important examples of symmetric matrices in multivariate statistics include the variance-covariance matrix and the correlation matrix.

## Linear Combination of Two Matrices

- Two matrices may be added if and only if they have the same dimensions. To add two matrices, add the corresponding elements.

```{r}
y <- as.matrix(baby[5:7,1:2])
y
x+y
```
 
- If ${\bf A}$ and ${\bf B}$ are two $n \times m$ dimensional matrices, $\alpha$ and $\beta$ are two scalars, then ${\bf C} = \alpha \, {\bf A} + \beta \, {\bf B} \in {\mathbb R}^{n \times m}$. A scalar multiples each entry of the matrix, i.e., 
\[
 {\bf C}_{i,j} = \alpha \, {\bf A}_{i,j} + \beta \, {\bf B}_{i,j}.
\]
 
```{r}
2*x + 3*y
```

## Multiplication of Two Matrices

- The **dot product** of two vectors $\underline{x}$ and $\underline{y}$ is the sum of the product of their entries: $\underline{x} \cdot \underline{y} = \sum_{i=1}^n x_i \, y_i$.

```{r}
sum(x[,1]*x[,2])
t(x[,1,drop=FALSE]) %*% x[,2,drop=FALSE]
```

- We multiply matrices by taking dot products of row and column vectors: ${\bf C} = {\bf A} \, {\bf B}$ is defined if number of columns in ${\bf A}$ equals the number of rows in ${\bf B}$. Then 
\[
 {\bf C}_{i,j} = {\bf A}_{i, -} \cdot {\bf B}_{-,j} = \sum_{k=1}^p {\bf A}_{i,k} \, {\bf B}_{k,j}.
\]
Here ${\bf A} \in {\mathbb R}^{n \times p}$, ${\bf B} \in {\mathbb R}^{p \times m}$, and ${\bf C} \in {\mathbb R}^{n \times m}$.

```{r}
x %*% t(y)
sum(x[1,]*t(y)[,3])
```

## Identity Matrix

- Definition: An identity matrix ${\bf I}$ is a square matrix that has ones on its diagonal (from upper left to bottom right) and has zeros elsewhere.

```{r}
eye <- diag(2)
eye
```

- When multiplying any square matrix, it leaves the matrix unchanged: ${\bf I} \, {\bf A} = {\bf A}$.

```{r}
z <- as.matrix(baby[1:2,1:2])
z
eye %*% z
```

## Matrix Inverse  

- Definition: The inverse of a square matrix ${\bf A}$, if it exists (the inverse of zero does not exist!?), is denoted by ${\bf A}^{-1}$, and satisfies
\[
 {\bf A}^{-1} \, {\bf A} = {\bf A} \, {\bf A}^{-1} = {\bf I}.
\]

```{r}
a <- t(x) %*% x
a_inv <- solve(a)
a_inv
a %*% a_inv
a_inv %*% a
```

*What is a numerical zero in R?*

- If the inverse exists, we say the matrix is **invertible**.

## Matrix Trace

- Definition: The trace of an $n \times n$ matrix ${\bf A}$ is the sum of its diagonal elements: 
\[
  \mbox{tr} {\bf A} = \sum_{i=1}^n {\bf A}_{i,i}.
\]

```{r}
sum(diag(a))
```

- For two matrices ${\bf A} \in {\mathbb R}^{n \times p}$, ${\bf B} \in {\mathbb R}^{p \times n}$, 
\[
 \mbox{tr} {\bf A} \, {\bf B} = \mbox{tr} {\bf B} \, {\bf A}.
\]

```{r}
x
z <- t(y)
z
sum(diag(x %*% z))
sum(diag(z %*% x))
```

## Eigenvalues and Eigenvectors

- Given a symmetric matrix ${\bf A}$, then any scalar $\lambda$ and vector $\underline{v}$ such that
\[
 {\bf A} \, \underline{v} = \lambda \, \underline{v}
\]
 are an **eigenvalue** and **eigenvector** (together called an **eigenpair**).
 
```{r}
eig <- eigen(a)
a %*% eig$vector[,1]
eig$values[1] * eig$vector[,1]

a %*% eig$vector[,2]
eig$values[2] * eig$vector[,2] 
```
  .
- A $n \times n$ symmetric matrix has $n$ real eigenvalues, denoted $\lambda_1, \ldots, \lambda_n$.

- Fact: $\mbox{tr} {\bf A} = \sum_{i=1}^n \lambda_i$.

```{r}
eig$values
sum(eig$values)
sum(diag(a))
```

- Fact: the determinant is the product of the eigenvalues: $\mbox{det} {\bf A} = \prod_{i=1}^n \lambda_i$.

## Positive Definite Matrix

- A symmetric $n \times n$ dimensional matrix ${\bf A}$ is **positive definite** if $\underline{v}^{\prime} {\bf A} \underline{v} > 0$ for all non-zero length $n$ vectors $\underline{v}$.  Compactly written ${\bf A} > 0$.

```{r}
v <- rnorm(2)
t(v) %*% a %*% v
```
 
- A symmetric $n \times n$ dimensional matrix ${\bf A}$ is **positive semi-definite** if $\underline{v}^{\prime} {\bf A} \underline{v} \geq 0$ for all length $n$ vectors $\underline{v}$.  Compactly written ${\bf A} \geq 0$.

- Fact: a symmetric $n \times n$ dimensional matrix is **positive definite** if and only if all of its eigenvalues are positive. (Remember, they must be real by symmetry!)

- Fact: a **positive definite** matrix is invertible.
